---
title: "Geography 13"
author: "[Oceane Roux](https://oceanerouxx.github.io/github.io/)"
subtitle: 'Lab 03: Distances and the Border Zone'
output:
  html_document:
    theme: journal
---

```{r setup, include=FALSE}
# SPDS
library(tidyverse)
library(sf)
library(units)

# Data
library(USAboundaries)
library(rnaturalearth)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

# Question 1

## 1.1 Define a Projection
```{r projection}
eqdc = '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

## 1.2 Get USA state boundaries
```{r}
conus= USAboundaries::us_states(resolution = "low") %>% 
  filter(!name %in% c("Hawaii", "Alaska", "Puerto Rico")) %>% 
  st_transform(eqdc)
```

## 1.3 County Boundaries
```{r}
countries= rnaturalearth::countries110 %>% 
  st_as_sf(countries) %>%  
  filter(admin %in% c("United States of America", "Mexico", "Canada")) %>% 
  st_transform(eqdc)
```

## 1.4 Get city locations from the CSV file
```{r, message=FALSE}
url = 'https://raw.githubusercontent.com/mikejohnson51/exercise-04/master/data/uscities.csv'
us_cities = readr::read_csv(url) %>% 
  filter(!state_name %in% c("Hawaii", "Puerto Rico", "Alaska")) %>% 
  st_as_sf(coords= c("lng", "lat"), crs= 4326) %>% 
  st_transform(eqdc)
```

# Question 2

## 2.1 Distance to USA Border (km)
```{r, echo=FALSE}
conus = USAboundaries::us_states() %>%
  filter(!state_name %in% c("Puerto Rico",
                            "Alaska",
                            "Hawaii"))

conus <- st_transform(conus, eqdc)

us_border = st_union(conus) %>%
    st_cast("MULTILINESTRING")
plot(us_border)

border_distances= us_cities %>% 
  mutate(distance_border= st_distance(., us_border),
  distance_border= units::set_units(distance_border, "km"),
  distance_border= as.numeric(distance_border))

table2.1= border_distances %>% 
  select("city", "state_name", "distance_border") %>% 
  slice_max(distance_border, n= 5) %>%
  st_drop_geometry()

  
knitr::kable(table2.1,
              caption= "Farthest Cities from USA Border",
              col.names= c( "City", "State", "Distance (km)"),
             format.args = list(big.mark = ",")) 
              

```


## 2.2 Distance to States (km)
```{r, echo=FALSE, warning=FALSE}
conus = USAboundaries::us_states() %>%
  filter(!state_name %in% c("Puerto Rico",
                            "Alaska",
                            "Hawaii")) 
conus <- st_transform(conus, eqdc)


state_border=  st_combine(conus) %>%
    st_cast("MULTILINESTRING")

plot(state_border)

border_distances_state= us_cities %>% 
  mutate(statedistance_border= st_distance(., state_border),
  statedistance_border= units::set_units(statedistance_border, "km"),
  statedistance_border= as.numeric(statedistance_border))

table2.2= border_distances_state %>% 
  select( "city", "state_name","statedistance_border") %>% 
  slice_max(statedistance_border, n= 5) %>%
  st_drop_geometry()

knitr::kable(table2.2,
              caption= "Farthest Cities to State Borders",
              col.names= c("City", "State", "Distance (km)"),
              format.args= list(big.mark= ".")) 

```

## 2.3 Distance to Mexico (km)
```{r, echo=FALSE, warning=FALSE}
mex= countries %>% 
  filter(admin %in% 'Mexico')
  
 mex_border= st_union(mex) %>%
    st_cast("MULTILINESTRING")
plot(mex_border)

mex_distance= us_cities %>% 
  mutate(mex_dis_border=st_distance(., mex),
         mex_dis_border=units::set_units(mex_dis_border, "km"),
  mex_dis_border= as.numeric(mex_dis_border))

table2.3= mex_distance %>% 
  select("city", "state_name", "mex_dis_border") %>% 
  slice_max(mex_dis_border, n=5) %>%
  st_drop_geometry()

knitr::kable(table2.3,
              caption= "Top 5 Furthest Cities from Mexico Border",
              col.names= c("City", "State",  "Distance (km)"),
              format.args= list(big.mark= "."))

```

## 2.4 Distance to Canada (km)
```{r, echo=FALSE, message=FALSE, warning=FALSE}
canada= countries %>% 
  filter(admin %in% 'Canada') 

(canada_border = st_union(canada) %>%
  st_cast("MULTILINESTRING"))

plot(canada_border)

canada_distance= us_cities %>% 
  mutate(can_dis=st_distance(us_cities, canada),
         can_dis=units::set_units(can_dis, "km"),
  can_dis= as.numeric(can_dis))

table2.4= canada_distance %>% 
  select("city","state_name","can_dis") %>% 
  slice_max(can_dis, n=5) %>%
  st_drop_geometry()

knitr::kable(table2.4,
              caption= "Top 10 Furthest Cities from Canada Border",
              col.names= c("City", "State Name", "Distance (km)"),
              format.args= list(big.mark= "."))  

```

# Question 3

## 3.1 Data
```{r, echo=FALSE}
borders = st_combine(countries) %>%
    st_cast("MULTILINESTRING")

state_borders=  st_combine(conus) %>%
    st_cast("MULTILINESTRING")

lrg_cities= us_cities %>% 
  slice_max(population, n= 10)

ggplot() + 
  geom_sf(data = borders, aes(), size = .1) + 
  geom_sf(data = state_borders, aes(), col = "dark red")+
  geom_sf(data = lrg_cities, aes(), col=" blue") + 
  ggrepel::geom_label_repel(data = lrg_cities,
                            aes(label = city, geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3) +
  labs(title = "Cities with Largest Populations",
       subtitle = "United States",
       x = "Longitude",
       y = "Latitude")  
  
```

## 3.2 Distance from the Border
```{r, echo=FALSE}

cities_border= border_distances %>% 
  slice_max(distance_border, n=5)

ggplot() + 
  geom_sf(data = us_border, aes())+
  geom_sf(data = border_distances, aes(col = distance_border), size = 0.3) +
  scale_color_gradient(low= "green4", high= "red")+
  geom_sf(data=cities_border, aes(), col= "blue")+
  geom_sf(data = cities_border, aes()) + 
  ggrepel::geom_label_repel(data = cities_border,
                            aes(label = city, geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3) +
  labs(title = "Cities Farthest from US Border",
       col= "Distance in km",
       x = "Longitude",
       y = "Latitude") 
```


## 3.3 City Distance from Nearest State
```{r, echo=FALSE}
cities_state = slice_max(border_distances_state, statedistance_border, n=5)

ggplot() + 
  geom_sf(data = conus, col = "black") +
  geom_sf(data = border_distances_state, aes(col= statedistance_border), size = 0.3) +
  geom_sf(data = cities_state, col = "blue") +
  scale_color_gradient(low= "green4", high= "red")+
  geom_sf(data = state_border, aes()) + 
  ggrepel::geom_label_repel(data = cities_state,
                            aes(label = city, geometry = geometry),
                            stat = "sf_coordinates",
                            size = 3) +
  labs(title = "Cities Farthest from State Borders",
       col= "Distance in km",
        x = "Longitude",
       y = "Latitude") 
```

## 3.4 Equidistance Boudary from Mexico to Canada
```{r, echo=FALSE, warning=FALSE}
EQ_cities = st_join(mex_distance, canada_distance, by = "city") %>%
  mutate(CM_cities = abs(can_dis - mex_dis_border))

equidistant = EQ_cities %>% 
  filter(CM_cities <= 100)

EQ_top5 = slice_max(equidistant, population.x, n = 5)

ggplot() + 
  geom_sf(data = state_border, col = "black") + 
  geom_sf(data = EQ_cities, col = "red3", size = 0.3) +
  geom_sf(data = EQ_top5, col = "blue") +
  gghighlight(CM_cities <= 100) +
  ggrepel::geom_label_repel(data = EQ_top5,
                            aes(label = city.x, geometry = geometry), 
                            stat = "sf_coordinates",
                            size = 3) +
  labs(title = "US Cities Equidistant to Mexico and Canada",
       x = "Longitude",
       y = "Latitude")
```

# Question 4

# 4.1 - Quantifying border zone
```{r, echo=FALSE}

cities_dist = border_distances %>% 
  filter(distance_border <= 160)

cities_dist_2 = length(cities_dist$city)

pop_cities_dist = sum(cities_dist$population)

pop_cities = sum(us_cities$population)

percent_pop_100 = pop_cities_dist/ pop_cities * 100

knitr::kable(tibble(cities_dist_2, pop_cities_dist, percent_pop_100), caption = "Statistics of 100 Miles of the US Border",
             col.names = c("Cities within 100 Miles Zone",
                           "Total Population within 100 Mile Zone",
                           "Percentage of Population within 100 Mile Zone"))
```
# 4.2 - Mapping border zone
```{r, echo=FALSE, warning=FALSE}

top10 = slice_max(cities_dist, population, n = 10)

ggplot() +
  geom_sf(data = conus, color = "black") +
  geom_sf(data = border_distances, aes(color = distance_border), size = 0.3) +
  geom_sf(data = top10, color = "grey") +
  scale_color_gradient(low = "orange", high = "darkred") +
  gghighlight(distance_border <= 160) +
  ggthemes::theme_map() +
  ggrepel::geom_label_repel(
    data = top10,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "Top 10 Most Populous Cities Within 100mi Danger Zone",
       color = "Distance (km)")
```




