---
title: "Geography 13"
author: "[Oceane Roux](https://oceanerouxx.github.io/github.io/)"
subtitle: 'Lab 02: COVID-19 Pandemic'
output:
  html_document:
    theme: journal
---


``` {r, include=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(readxl)
library(zoo)

```


# Data

``` {r, results="hide"}
url = 'https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv'
covid = read_csv(url)

landDate = read_csv("C:/Users/rouxo/Documents/github/geog-13-labs/data/landdata-states.csv")
popEst = read_xls("C:/Users/rouxo/Documents/github/geog-13-labs/data/PopulationEstimates.xls", skip=2)
```

# Question 1

## Top California Counties with the Most Cumulative Cases
``` {r, echo=FALSE}
state.interest = "California"

CAcumulative = covid %>% 
  filter(state== state.interest, date == max(date)) %>%
  slice_max(cases, n = 5) %>%
  select(county, cases)

kable(CAcumulative, col.names = c("County", "Cases"), 
               caption = "Cumulative Cases",
               format.args = list(big.mark = ","))

```

## Top California Counties with the Most New Cases
```{r, echo=FALSE}
state.interest = "California"

CAnewcases = covid %>% 
  filter(state== state.interest) %>%
  group_by(county) %>% 
  mutate(dailyNewCases = cases - lag(cases)) %>%
  ungroup() %>% 
  filter(date == max(date)) %>%
  slice_max(dailyNewCases, n = 5) %>%
  select(county, dailyNewCases)

kable(CAnewcases, col.names = c("County", "Cases"),
      caption = "New Cases",
      format.args = list(big.mark = ","))

```

## Joining Population Data and California COVID Data
```{r}
state.interest = "California"

pop = popEst %>%
  select(Area_Name, POP_ESTIMATE_2019, FIPStxt) %>%
  rename(fips = FIPStxt)

newcovid = covid %>% 
  filter(state== state.interest) %>%
  group_by(county) %>% 
  mutate(dailyNewCases = cases - lag(cases)) %>%
  ungroup() %>% 
  filter(date == max(date))

covidpop = inner_join(newcovid, pop, by = "fips") %>%
  filter(state == state.interest)

```

## Most Cumulative Cases per Capita
```{r, echo=FALSE}
state.interest = "California"

cumcapita = covidpop %>%
  filter(state == state.interest, date == max(date)) %>%
  mutate(perCapita = cases/POP_ESTIMATE_2019) %>%
  slice_max(perCapita, n = 5) %>%
  select(county, perCapita)

kable(cumcapita, col.names = c("County", "Cases per Capita"),
      caption = "Cases per Capita",
      format.args = list(big.mark = ","))

```

## Most New Cases per Capita
```{r, echo=FALSE}
state.interest = "California"

newcapita = covidpop %>%
  filter(state == state.interest, date == max(date)) %>%
  mutate(dailyNewCases = cases - lag(cases)) %>%
  mutate(perCapita = dailyNewCases/POP_ESTIMATE_2019) %>%
  slice_max(perCapita, n = 5) %>%
  select(county, perCapita)

kable(newcapita, col.names = c("County", "Cases per Capita"),
      caption = "New Cases per Capita",
      format.args = list(big.mark = ","))

```

## Average of New Cases in the Last 14 Days per 100,000 People
```{r, echo=FALSE, message=FALSE}
state.interest = "California"

avg_14 = covidpop %>%
  filter(state == state.interest, date > max(date) - 14) %>%
  group_by(county, state) %>% 
  summarise(cases100k = sum(dailyNewCases, na.rm = TRUE)/(mean(POP_ESTIMATE_2019, na.rm = TRUE)*100000)) %>%
  ungroup() %>%
  group_by(state) %>%
  summarise(total_cases_14 = sum(cases100k / 14))

kable(avg_14, col.names = c("State", "Cases per 100k"),
      caption = "New Cases per 100,00 People",
      format.args = list(big.mark = ","),
      digits = 9)
```

```{r}
state.interest = "California"

totalcases = covid %>% 
  filter(state== state.interest, date == max(date)) %>%
  summarise(sum(cases))
  
totalnew = covid %>% 
  filter(state== state.interest) %>%
  group_by(county) %>% 
  mutate(dailyNewCases = cases - lag(cases)) %>%
  ungroup() %>% 
  filter(date == max(date)) %>%
  summarise(sum(dailyNewCases))

safe = covid %>%
  filter(state== state.interest) %>%
  group_by(county) %>% 
  mutate(dailyNewCases = cases - lag(cases)) %>%
  ungroup() %>% 
  filter(date == max(date)) %>%
  filter(dailyNewCases == "0")
  
```

### Total cases of COVID in California: 3,839,640
### Total new cases of COVID in California: 4131
### Safe counties is California: 19
 

# Question 2

## 7 Day Rolling Mean of COVID Cases
``` {r, echo=FALSE, message=FALSE, warning=FALSE}
states.of.interest <- c("New York", "California", "Louisiana", "Florida")

StateLevel = covid %>%
  filter(state %in% states.of.interest) %>%
  group_by(state, date) %>%
  summarise(cases = sum(cases, nar.rm=TRUE)) %>%
  mutate(dailyNewCases = cases -lag(cases), RollMean7 = rollmean(dailyNewCases, 7, fill = NA, align = "right"))

ggplot(data = StateLevel, aes(x = date, y = RollMean7, group=state)) +
  geom_line(aes(col=state)) +
  labs(x="Date", y="Daily New Cases", title="Daily New COVID Cases by State ", subtitle = "Data from NY Times") +
  facet_wrap(~state) +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))
  

```


